---
title: "Lecture1"
output: rmarkdown::github_document
---

* 2025-03-11


### Computing on a data matrix

```{r}
gene1 = c(1.00, 1.50, 1.25)
gene1
```

```{r}
gene2 = c(1.35, 1.55, 1.00)
gene3 = c(-1.10, -1.50, -1.25)
gene4 = c(-1.20, -1.30, -1.00)
genes = c(gene1, gene2, gene3, gene4)
genes
```

```{r}
geneData = matrix(genes, nrow=4, ncol=3, byrow=TRUE) # !byrow!
geneData
```

```{r}
rownames(geneData) = c("gene1", "gene2", "gene3", "gene4")
colnames(geneData) = c("Eric", "Peter", "Anna")
geneData
# t(geneData)
```

```{r}
# data를 받고 제일 처음 해야할 것.
nrow(geneData) # # of gene
ncol(geneData) # # of sample
class(geneData)
```

```{r}
geneData[3,] # third row (gene) data
geneData[c(1,2), ] # first and second row(gene) data
geneData[c("gene1", "gene2"), ] # can indexing with names
```

```{r}
apply(geneData, 1, mean) # 1 means mean of each row(gene) == row by row
apply(geneData, 2, mean) # 2 means mean of each column(sample) == col by col
```

### Constructing a data frame

* Matrix must have same data type init.
* But Data frame do not need it
* 하나의 커럼에서는 똑같은 tpye이 들어가야함.
* 각 행의 열 수 가 같아야함.

```{r}
patients.df = data.frame(
  patientID = c("101", "102", "103", "104"),
  treatment = c("drug", "placebo", "drug", "placebo"),
  age = c(20, 30, 24, 22) 
)

patients.df
```

```{r}
nrow(patients.df)
ncol(patients.df)
class(patients.df)
```



### Referencing column vectors in a data frame

```{r}
patients.df[, 2]

patients.df[[2]]
patients.df[["treatment"]]
patients.df$treatment # 이걸 제일 많이 사용함
patients.df[, 2]
```

```{r}
patients.df[c(1,3), ]

patients.df$treatment=="drug"
patients.df[patients.df$treatment=="drug", ] # [, (모든 열)]

subset(patients.df, treatment=="drug")
```

```{r}
row.names(patients.df) = patients.df$patientID
patients.df[c("101", "103"),]
```


```{r}
patients.df[2] # series
patients.df[[2]] # values

patients.df["treatment"]
patients.df[["treatment"]]
```

### Constructing a list

* list는 매우 자유로움!

```{r}
list1 <- list(
  c("p53", "p63", "p73"),
  matrix(rnorm(10), nrow=2),
  c(TRUE, FALSE, TRUE, FALSE, FALSE)
)
list1
```

```{r}
p53FamilyGenes <- c("p53", "p63", "p73")
list2 <- list(
  genes = p53FamilyGenes,
  geneExpression = matrix(rnorm(10), nrow=2),
  remission = c(TRUE, FALSE, TRUE, FALSE, FALSE)
)
list2
```

```{r}
list2[1]
list2[[1]]
list2[["genes"]]
list2$genes
```


```{r}
list2[[1]][1] <- "her2"
list2$genes
list2
```


```{r}
list2[[2]]
list2$geneExpression
```


```{r}
list2[[3]]
list2$remission
```


### Data path attachment

```{r}
# remission # error
attach(list2) # 데이터 이름 생략하기 위해서..!
remission

detach(list2)
# remission # error
```


```{r}
genes <- c("p53", "p63", "p73")
attach(list2)
genes # 우선순위는 기존 객체..
list2$genes
```


```{r}
patients.df$treatment
attach(patients.df)

treatment
detach(patients.df)
```

### Generating factor

* For plotting and testing of hypo we need to generate a **factor**.

```{r}
gl(3, 5) # gl( # of group, # of element of each group) # factor
rep(1:3, each = 5) # vector
```

* factor type은 범주형 자료라 연산 안됨. (확인해보기)

```{r}
class(gl(3, 5))
class(rep(1:3, each = 5))
```


```{r}
factor(rep(1:3, each=5)) # same as gl()

A <- factor(rep(1:3, each = 5))
class(A)
```


### Functions

```{r}
set.seed(123)

A <- matrix(sample(1:100, 12), nrow=3, ncol=4)

A

apply(A, 2, var)
```

```{r}
fun <- function(x) {
  sum((x-mean(x))^2)/(length(x)-1)
}

apply(A, 2, fun)

apply(A, 2, function(x) sum((x-mean(x))^2)/(length(x) - 1))
```

* case : more than one argment

```{r}
fun1 <- function(x, a) sum(x-a)^2/(length(x)-1)
apply(A, 2, fun1, a=40)
```


```{r}
fun2 <- function(x, a, b) sum(x - a)^2 / b
apply(A, 2, fun2, a = 40, b = 10)
```

```{r}
list(a=2, b=3)

lapply(list(a=2, b=3), sqrt) # 각각의 원소마다 적용하라..

matrix(1:6, 2, 3)
lapply(matrix(1:6, 2, 3), function(x) x^2)
```

```{r}
mylapply <- function(lst, fun) {
  lst <- as.list(lst)
  ans <- vector("list", length(lst))
  names(ans) <- names(lst)
  for(i in 1:length(lst))
    ans[i] <- list(fun(lst[[i]]))
  ans
  }
mylapply(list(a=2, b = 3), sqrt)
```

### Introduction to Bioconductor

```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")
#BiocManager::install()
```


```{r}
#BiocManager::install("ALL")
library(ALL)
```

```{r}
data(ALL)
#?ALL
ALL
str(ALL)
```


```{r}
class(ALL)
```

```{r}
experimentData(ALL)
#sampleNames(ALL)
#featureNames(ALL)
```

```{r}
dim(exprs(ALL))
exprs(ALL)[1:10, 1:5]
```

```{r}
varLabels(ALL)
ALL$sex
ALL$age
ALL$mol.biol
```

### ALL Data Example

```{r}
ALL$BT

table(ALL$BT)
```

```{r}
BT <- ALL$BT %in% c("T", "T1", "T2", "T3", "T4")
y <- factor(as.numeric(BT), labels=c("B", "T")) # logical로 처리
table(y)
```

```{r}
x <- t(exprs(ALL))
dim(x)
summary(as.numeric(x)) # 이거 체크 vs summary(x)

hist(x, col="orange", nclass=50, main="Expression Data")
```


* scaling 할 때 행은 sample 열은 variable로 설정하기!!!!!!!

```{r}
x2 <- scale(x)
dim(x2)
summary(as.numeric(x2))
hist(x2, col="orange", nclass=50, main="Scaled Expression Data")
```





























